---
title: "Augmentation file"
format: html
editor: visual
---

```{r}
#|label: load the packages
library(tidyverse)
library(stringr)
library(dplyr)
library(tidyr)
```

Here we begin by loading our two clean mouse datasets; the main one and the supplemantary data

```{r}
Sys.setenv(VROOM_CONNECTION_SIZE = 3062144) # Setting a larger connection buffer to be able to load the data
# Loading the data
mouse_data <- read_tsv(file = "../data/02_dat_clean_mouse.tsv.gz")
mouse_data_suppl <- read_tsv(file = "../data/02_dat_clean_mouse_suppl.tsv.gz")
```

Now we add the metadata to the main data

```{r}
#Join main and metadata
mouse_df <- left_join(mouse_data, mouse_data_suppl, join_by(treatment))
```

We now create a new df where we lose the control group experiments (DMSO) and instead divide the noin-control values with the DMSO values of each replicate from the other treatment gene expressions, and thereby convert from expression value to relative expressions. We then log2 transofrm the relative expression, such that a 0 means no change between control and treatment, and positive values mean upregulated and negative means down_regulated $$
\log(\frac{expression\_control\_trial (DMSO)}{expression\_drugX\_trial})
$$

This operation will be performed now

```{r}
"Make a baseline dataset with the values for the control group"
DMSO <- mouse_df |> 
  filter(treatment=='DMSO') |> 
  select(-c(treatment))

"Making the relative dataset"
rel_mouse_df <- mouse_df |> 
  left_join(DMSO,                    # Creating baseline columns for each gene with the baseline values for each replicate 
            by = 'replicate', 
            suffix = c("", "_baseline")
            ) |> 
  mutate(across(-c("treatment",      # Subtracting baseline values from the original gene columns
                   "replicate", 
                   `Factor Value[compound]`, 
                   ends_with('baseline')
                   ),
                ~ .x - get(paste0(cur_column(), "_baseline"))
                )
         ) |>         # Replace NaNs with 0 since R returns 0 for log(0/0,2)
  select(-c(ends_with('baseline')))  # loose the baseline helper columns


```


```{r}
write_tsv(x = rel_mouse_df, file = str_c("../data/04_dat_aug_mouse", ".tsv.gz"))
```
