---
title: "Clean file"
format: html
editor: visual
---

```{r}
#|label: load the packages

library(tidyverse)
library(stringr)
library(dplyr)
library(tidyr)
```

Now we need to clean our data First we will check the structure of the data and convert them into tibble (better performance for large datasets)

```{r}
#| label: load as tibbles and data structure
human_data <- vroom::vroom("../data/human.tsv", delim = "/t")
human_sup_data <- vroom::vroom("../data/human_sup.tsv", delim = "/t")

glimpse(human_data)
glimpse(human_sup_data)
head(human_data)
head(human_sup_data)
```

```{r}
# Test for missing data
if (any(is.na(mouse_data))) {
  cat("There are NA values in the data frame.\n")
} else {
  cat("There are no NA values in the data frame.\n")
}
```

From this illustration we can see that there are no na data detected in our human raw dataframe

Next we will rename the column names as treated and untreated based on DMSO (vehicle solvent for the drug) and G007-LK (drug) respectively. We observe that there are many variable in column names, so we need to separate them

**we need to see the unit of these expressions levels**

```{r}
library(dplyr)

# Sample tibble
my_tibble <- tibble(
  old_col1 = c(1, 2, 3),
  old_col2 = c(4, 5, 6),
  old_col3 = c(7, 8, 9)
)

my_tibble

# Function to modify column names
modify_column_names <- function(name) {
  # Replace "old_" with "new_"
  gsub("old_", "new_", name)
}

# Use rename_with to apply the custom function to column names
my_tibble <- my_tibble %>%
  rename_with(~modify_column_names(.), starts_with("old_col"))

# Print the updated tibble
print(my_tibble)
```

```{r}
#|label: rename and separate columns names

#Maybe the followed code should be separated in two chunks
human_data_clean  <- human_data |> 
   treatment_cols(across(everything()))
  
#   rename_with(
#     ~case_when(
#       str_detect(.,".DMSO") ~ str_replace_all(.,
#                                             ".DMSO",
#                                             "_untreated"),
#       str_detect(.,".G007.LK") ~ str_replace_all(.,
#                                                ".G007.LK",
#                                                "_treated"),
#       TRUE ~ .  # If no pattern is matched, keep the original name
#     ),
#     everything()# Apply case when function in each column
#   )
# 
# head(human_raw_data_clean)
```

```{r}
human_raw_data_clean |>
  group_by(symbol) |>
  summarise(n = n()) |> 
  filter(n> 1)

#there are 1539 genes that appeared in more than one geneId 
```

```{r}
#Pivot the human df
human_raw_data_clean_long<- human_raw_data_clean |> 
  pivot_longer(
    cols = !(gene_id:symbol),
    names_to = c("cell_line", 
                 "treatment"), 
    names_sep = "_",
    values_to = "expression_levels",
    values_drop_na = TRUE
      ) 

head(human_raw_data_clean_long)
```

```{r}
#| label: This one goes to the describe file
#check the dataset for duplicates
human_raw_data |>
  group_by(symbol) |> 
  add_count(symbol) |> 
  filter(n>1) |> 
  select(n)
#these results indicate presence of different isomorphs  
```

```{r}
#|label: group by genes
# not ready yet
human_raw_data_clean |> 
  select(-gene_id) |> 
  group_by(symbol) |> 
  nest() |> 
  head()
```

```{r}

#|label: selection of the useful columns
# need to rename the columns
human_sup_data|> 
  as_tibble() |> 
  select(`Source Name`,`Characteristics[genotype]`)
```

**To do:**

check and manipulate possible NAs

discard gene ID and use only symbol for each

Fix the name in sup dataset

```{r}
colnames(human_raw_data_clean)
```

```{r}
human_raw_data_clean |>
  group_by(symbol) |>
  summarise(n = n()) |> 
  filter(n> 1)
```

```{r}
colnames(human_suppl_raw)
```

Supplementary files

```{r}
#Test for missing data
is.na(human_suppl_raw)
```

```{r}
colnames(human_suppl_raw)<- c("Name", "Organism","Developmental_stage", "Disease","Genotype","Organism_part",   "Cell_line","Material_Type","Protocol_REF","Protocol_REF1","Protocol_REF2","Performer", "Extract_Name",   "Library_Layout", "Library_Selection","Library_Source","Library_Strategy","Protocol_REF3", "Performer_1","Assay_Name","Technology_Type","Protocol_REF4", "Performer_2","Derived_Array_Data_File", "Derived_ArrayExpress_FTP_file", "Factor_Value_compound", "Factor_Value_dose", "Unit_dose_unit")
```

```{r}
#change the name of the columns of the human supplementary dataset
colnames(human_suppl_raw)
```

```{r}
head(human_suppl_raw)
```

```{r}
#separate the Name and the Cell_lines columns
human_suppl_raw |> 
  group_by(Name) |> 
  separate(Name, into = c("Cell_line1","Compound"), sep = "\\.")
```

```{r}
human_suppl_raw |> 
  distinct(Name)
```

```{r}
# create the vector of the feature Name values 
human_suppl_clean<-data.frame(Name = c(
    "A375.DMSO", "LOX.DMSO", "Skmel28.DMSO", "WM115.DMSO", "WM1341.DMSO",
    "WM239.DMSO", "WM266.DMSO", "WM35.DMSO", "WM45.1.DMSO", "WM793.DMSO",
    "WM9.DMSO", "WM983.DMSO", "MeWo.DMSO", "FemxI.DMSO", "FemxV.DMSO",
    "WM1366.DMSO", "WM852.DMSO", "WM1382.DMSO", "A375.G007.LK", "LOX.G007.LK",
    "Skmel28.G007.LK", "WM115.G007.LK", "WM1341.G007.LK", "WM239.G007.LK",
    "WM266.G007.LK", "WM35.G007.LK", "WM45.1.G007.LK", "WM793.G007.LK",
    "WM9.G007.LK", "WM983.G007.LK"
  )
)

# Now we want to split those values depending on the compound they used

compound_pattern <- c("DMSO", "G007.LK")


```

```{r}
human_suppl_clean |> 
  mutate(
    Cell_Line1 = sub(paste(compound_pattern, collapse = "|"), "", Name),
    column2 = sub(".*(" paste(compound_pattern, collapse = "|")").*", "\\1", Name)
  )
```
