---
title: "load file"
format: html
editor: visual
---

```{r}
#First we have to load the necessary libraries
library(tidyverse)
library(dplyr)
```

```{r}
#then we need to load our data and transform them to tsv format

setwd("C:/Users/antot/Desktop/master/R for Bio Data science/project/group_24_project")


#transform the human raw data
human_sup_txt<-vroom::vroom("_raw/human_suppl.txt", show_col_types = FALSE) #this contains the raw data
write.table(human_sup_txt, file = "data/human_sup.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

#transform the human metadata
human_data_csv<- vroom::vroom("_raw/human.csv", show_col_types = FALSE)
write.table(human_data_csv, file = "data/human_data.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

#transform the mouse raw data
mouse_sup_txt<-vroom::vroom("_raw/mouse_suppl.txt", show_col_types = FALSE)
write.table(mouse_sup_txt, file = "data/mouse_sup.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

#transform the mouse metadata
mouse_data_gz<-vroom::vroom("_raw/mouse.gz", show_col_types = FALSE)
write.table(mouse_data_gz, file = "data/mouse_data.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

```

Then we can take a first look on the structure of the datasets.

This dataset comes from the study (**Mouse melanoma cell line B16-F10 treated with G007-LK**) in which 12 samples were used. What was measured in that study was the effect of different compounds in the B16-F10 cells. The compounds that were used were the following: dimethyl sulfoxide, WNT3a and GOO7LK (Each of the compound was applied in equal number of samples ie. 3).

This study could be found under the Accession number: E-MTAB-8101

The dataset containing the human cells was generated though the **Human melanoma cell lines treated with G007-LK** study during which 18 human melanoma cell lines derived from the skin of the donors were treated with the compound GOO7-LK with the purpose of detecting its effect to the cells and in order to compare those results with a similar experiment performed on mice cell lines.

In detail the experiment went like so:

Three pooled technical replicates from a panel of 18 human melanoma cell lines were treated with DMSO and G007-LK for 24 hours.

This study could be found under the Accession number: E-MTAB-8438

```{r}
setwd("C:/Users/antot/Desktop/master/R for Bio Data science/project/group_24_project")

#we are going to load the tranformed datasets
mouse_raw<- vroom::vroom("data/mouse_sup.tsv", show_col_types = FALSE)
mouse_metadata<-vroom::vroom("data/mouse_data.tsv", show_col_types = FALSE)


human_raw<- vroom::vroom("data/human_sup.tsv", show_col_types = FALSE)
human_metadata<- vroom::vroom("data/human_data.tsv", show_col_types = FALSE)
```

```{r}
head(mouse_raw)
```

```{r}

head(mouse_metadata)
```

```{r}
head(human_raw)
```

```{r}
head(human_metadata)
```
